---
title: "Day04_Secure_Container_AOC_2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tictoc)
```

# Day 4: Secure Container

## Puzzle 1

You arrive at the Venus fuel depot only to discover it's protected by a password. The Elves had written the password on a sticky note, but someone threw it out.

However, they do remember a few key facts about the password:

- It is a six-digit number.
- The value is within the range given in your puzzle input.
- Two adjacent digits are the same (like `22` in `122345`).
- Going from left to right, the digits never decrease; they only ever increase or stay the same (like `111123` or `135679`).

Other than the range rule, the following are true:

`111111` meets these criteria (double 11, never decreases).
`223450` does not meet these criteria (decreasing pair of digits 50).
`123789` does not meet these criteria (no double).

How many different passwords within the range given in your puzzle input meet these criteria?

Your puzzle input is `108457-562041`

```{r}
input <- as.character(108457:562041)
```


```{r puzzle1}

tic("puzzle 1")

# at least two adjacent digits are the same
input <- input[str_detect(input, "(\\d)\\1")]

# numbers can never decrease
index1 <- 
  map_lgl(.x = input, 
          .f = ~ str_split(.x, "")[[1]] %>% 
            parse_integer() %>% 
            {. >= lag(., default = 0L)} %>% 
            {all(. == TRUE)} )

answer1 <- input[index1]

length(answer1)

toc()

```

`2779` unique possible passwords that match criteria.

## Puzzle 2

An Elf just remembered one more important detail: the two adjacent matching digits are not part of a larger group of matching digits.

Given this additional criterion, but still ignoring the range rule, the following are now true:

`112233` meets these criteria because the digits never decrease and all repeated digits are exactly two digits long.

`123444` no longer meets the criteria (the repeated `44` is part of a larger group of `444`).

`111122` meets the criteria (even though 1 is repeated more than twice, it still contains a double `22`).

How many different passwords within the range given in your puzzle input meet all of the criteria?

```{r puzzle2}

tic("puzzle 2")

index2 <- 
  map(.x = answer1, .f = ~ str_split(.x, "")[[1]]) %>% 
  map(table) %>% 
  map(~ any(.x == 2)) %>% 
  unlist()

answer2 <- answer1[index2]

length(answer2)

toc()

```

With this new criteria, `1972` passwords could work