---
title: "Day12_The_N-Body_Problem_AOC_2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 20) # scientific notion setting change

library(tidyverse)
```

# Day 12: The N-Body Problem

# Puzzle 1

The space near Jupiter is not a very safe place; you need to be careful of a big distracting red spot, extreme radiation, and a whole lot of moons swirling around. You decide to start by tracking the four largest moons: Io, Europa, Ganymede, and Callisto.

After a brief scan, you calculate the *position of each moon* (your puzzle input). You just need to *simulate their motion* so you can avoid them.

Each moon has a 3-dimensional position (x, y, and z) and a 3-dimensional velocity. The position of each moon is given in your scan; the x, y, and z velocity of each moon starts at 0.

Simulate the motion of the moons in *time steps*. Within each time step, first update the velocity of every moon by applying *gravity*. Then, once all moons' velocities have been updated, update the position of every moon by applying *velocity*. Time progresses by one step once all of the positions are updated.

To apply *gravity*, consider every *pair* of moons. On each axis (x, y, and z), the velocity of each moon changes by *exactly +1 or -1* to pull the moons together. For example, if Ganymede has an x position of 3, and Callisto has a x position of 5, then Ganymede's x velocity changes by +1 (because 5 > 3) and Callisto's x velocity changes by -1 (because 3 < 5). However, if the positions on a given axis are the same, the velocity on that axis does not change for that pair of moons.

Once all gravity has been applied, apply velocity: simply add the velocity of each moon to its own position. For example, if Europa has a position of x=1, y=2, z=3 and a velocity of x=-2, y=0,z=3, then its new position would be x=-1, y=2, z=6. This process does not modify the velocity of any moon.

For example, suppose your scan reveals the following positions:

`<x=-1, y=0, z=2>`
`<x=2, y=-10, z=-7>`
`<x=4, y=-8, z=8>`
`<x=3, y=5, z=-1>`

Simulating the motion of these moons would produce the following:

After 0 steps:
`pos=<x=-1, y=  0, z= 2>, vel=<x= 0, y= 0, z= 0>`
`pos=<x= 2, y=-10, z=-7>, vel=<x= 0, y= 0, z= 0>`
`pos=<x= 4, y= -8, z= 8>, vel=<x= 0, y= 0, z= 0>`
`pos=<x= 3, y=  5, z=-1>, vel=<x= 0, y= 0, z= 0>`

After 1 step:
`pos=<x= 2, y=-1, z= 1>, vel=<x= 3, y=-1, z=-1>`
`pos=<x= 3, y=-7, z=-4>, vel=<x= 1, y= 3, z= 3>`
`pos=<x= 1, y=-7, z= 5>, vel=<x=-3, y= 1, z=-3>`
`pos=<x= 2, y= 2, z= 0>, vel=<x=-1, y=-3, z= 1>`

After 2 steps:
`pos=<x= 5, y=-3, z=-1>, vel=<x= 3, y=-2, z=-2>`
`pos=<x= 1, y=-2, z= 2>, vel=<x=-2, y= 5, z= 6>`
`pos=<x= 1, y=-4, z=-1>, vel=<x= 0, y= 3, z=-6>`
`pos=<x= 1, y=-4, z= 2>, vel=<x=-1, y=-6, z= 2>`

After 3 steps:
`pos=<x= 5, y=-6, z=-1>, vel=<x= 0, y=-3, z= 0>`
`pos=<x= 0, y= 0, z= 6>, vel=<x=-1, y= 2, z= 4>`
`pos=<x= 2, y= 1, z=-5>, vel=<x= 1, y= 5, z=-4>`
`pos=<x= 1, y=-8, z= 2>, vel=<x= 0, y=-4, z= 0>`

After 4 steps:
`pos=<x= 2, y=-8, z= 0>, vel=<x=-3, y=-2, z= 1>`
`pos=<x= 2, y= 1, z= 7>, vel=<x= 2, y= 1, z= 1>`
`pos=<x= 2, y= 3, z=-6>, vel=<x= 0, y= 2, z=-1>`
`pos=<x= 2, y=-9, z= 1>, vel=<x= 1, y=-1, z=-1>`

After 5 steps:
`pos=<x=-1, y=-9, z= 2>, vel=<x=-3, y=-1, z= 2>`
`pos=<x= 4, y= 1, z= 5>, vel=<x= 2, y= 0, z=-2>`
`pos=<x= 2, y= 2, z=-4>, vel=<x= 0, y=-1, z= 2>`
`pos=<x= 3, y=-7, z=-1>, vel=<x= 1, y= 2, z=-2>`

After 6 steps:
`pos=<x=-1, y=-7, z= 3>, vel=<x= 0, y= 2, z= 1>`
`pos=<x= 3, y= 0, z= 0>, vel=<x=-1, y=-1, z=-5>`
`pos=<x= 3, y=-2, z= 1>, vel=<x= 1, y=-4, z= 5>`
`pos=<x= 3, y=-4, z=-2>, vel=<x= 0, y= 3, z=-1>`

After 7 steps:
`pos=<x= 2, y=-2, z= 1>, vel=<x= 3, y= 5, z=-2>`
`pos=<x= 1, y=-4, z=-4>, vel=<x=-2, y=-4, z=-4>`
`pos=<x= 3, y=-7, z= 5>, vel=<x= 0, y=-5, z= 4>`
`pos=<x= 2, y= 0, z= 0>, vel=<x=-1, y= 4, z= 2>`

After 8 steps:
`pos=<x= 5, y= 2, z=-2>, vel=<x= 3, y= 4, z=-3>`
`pos=<x= 2, y=-7, z=-5>, vel=<x= 1, y=-3, z=-1>`
`pos=<x= 0, y=-9, z= 6>, vel=<x=-3, y=-2, z= 1>`
`pos=<x= 1, y= 1, z= 3>, vel=<x=-1, y= 1, z= 3>`

After 9 steps:
`pos=<x= 5, y= 3, z=-4>, vel=<x= 0, y= 1, z=-2>`
`pos=<x= 2, y=-9, z=-3>, vel=<x= 0, y=-2, z= 2>`
`pos=<x= 0, y=-8, z= 4>, vel=<x= 0, y= 1, z=-2>`
`pos=<x= 1, y= 1, z= 5>, vel=<x= 0, y= 0, z= 2>`

After 10 steps:
`pos=<x= 2, y= 1, z=-3>, vel=<x=-3, y=-2, z= 1>`
`pos=<x= 1, y=-8, z= 0>, vel=<x=-1, y= 1, z= 3>`
`pos=<x= 3, y=-6, z= 1>, vel=<x= 3, y= 2, z=-3>`
`pos=<x= 2, y= 0, z= 4>, vel=<x= 1, y=-1, z=-1>`

Then, it might help to calculate the *total energy in the system*. The total energy for a single moon is its *potential energy multiplied by its kinetic energy*. A moon's potential energy is the sum of the absolute values of its x, y, and z position coordinates. A moon's kinetic energy is the sum of the absolute values of its velocity coordinates. Below, each line shows the calculations for a moon's potential energy (pot), kinetic energy (kin), and total energy:

Energy after 10 steps:
`pot: 2 + 1 + 3 =  6;   kin: 3 + 2 + 1 = 6;   total:  6 * 6 = 36`
`pot: 1 + 8 + 0 =  9;   kin: 1 + 1 + 3 = 5;   total:  9 * 5 = 45`
`pot: 3 + 6 + 1 = 10;   kin: 3 + 2 + 3 = 8;   total: 10 * 8 = 80`
`pot: 2 + 0 + 4 =  6;   kin: 1 + 1 + 1 = 3;   total:  6 * 3 = 18`
`Sum of total energy: 36 + 45 + 80 + 18 = 179`

In the above example, adding together the total energy for all moons after 10 steps produces the total energy in the system, 179.

Here's a second example:

`<x=-8, y=-10, z=0>`
`<x=5, y=5, z=10>`
`<x=2, y=-7, z=3>`
`<x=9, y=-8, z=-3>`

Every ten steps of simulation for 100 steps produces:

After 0 steps:
`pos=<x= -8, y=-10, z=  0>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  5, y=  5, z= 10>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  2, y= -7, z=  3>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  9, y= -8, z= -3>, vel=<x=  0, y=  0, z=  0>`

After 10 steps:
`pos=<x= -9, y=-10, z=  1>, vel=<x= -2, y= -2, z= -1>`
`pos=<x=  4, y= 10, z=  9>, vel=<x= -3, y=  7, z= -2>`
`pos=<x=  8, y=-10, z= -3>, vel=<x=  5, y= -1, z= -2>`
`pos=<x=  5, y=-10, z=  3>, vel=<x=  0, y= -4, z=  5>`

After 20 steps:
`pos=<x=-10, y=  3, z= -4>, vel=<x= -5, y=  2, z=  0>`
`pos=<x=  5, y=-25, z=  6>, vel=<x=  1, y=  1, z= -4>`
`pos=<x= 13, y=  1, z=  1>, vel=<x=  5, y= -2, z=  2>`
`pos=<x=  0, y=  1, z=  7>, vel=<x= -1, y= -1, z=  2>`

After 30 steps:
`pos=<x= 15, y= -6, z= -9>, vel=<x= -5, y=  4, z=  0>`
`pos=<x= -4, y=-11, z=  3>, vel=<x= -3, y=-10, z=  0>`
`pos=<x=  0, y= -1, z= 11>, vel=<x=  7, y=  4, z=  3>`
`pos=<x= -3, y= -2, z=  5>, vel=<x=  1, y=  2, z= -3>`

After 40 steps:
`pos=<x= 14, y=-12, z= -4>, vel=<x= 11, y=  3, z=  0>`
`pos=<x= -1, y= 18, z=  8>, vel=<x= -5, y=  2, z=  3>`
`pos=<x= -5, y=-14, z=  8>, vel=<x=  1, y= -2, z=  0>`
`pos=<x=  0, y=-12, z= -2>, vel=<x= -7, y= -3, z= -3>`

After 50 steps:
`pos=<x=-23, y=  4, z=  1>, vel=<x= -7, y= -1, z=  2>`
`pos=<x= 20, y=-31, z= 13>, vel=<x=  5, y=  3, z=  4>`
`pos=<x= -4, y=  6, z=  1>, vel=<x= -1, y=  1, z= -3>`
`pos=<x= 15, y=  1, z= -5>, vel=<x=  3, y= -3, z= -3>`

After 60 steps:
`pos=<x= 36, y=-10, z=  6>, vel=<x=  5, y=  0, z=  3>`
`pos=<x=-18, y= 10, z=  9>, vel=<x= -3, y= -7, z=  5>`
`pos=<x=  8, y=-12, z= -3>, vel=<x= -2, y=  1, z= -7>`
`pos=<x=-18, y= -8, z= -2>, vel=<x=  0, y=  6, z= -1>`

After 70 steps:
`pos=<x=-33, y= -6, z=  5>, vel=<x= -5, y= -4, z=  7>`
`pos=<x= 13, y= -9, z=  2>, vel=<x= -2, y= 11, z=  3>`
`pos=<x= 11, y= -8, z=  2>, vel=<x=  8, y= -6, z= -7>`
`pos=<x= 17, y=  3, z=  1>, vel=<x= -1, y= -1, z= -3>`

After 80 steps:
`pos=<x= 30, y= -8, z=  3>, vel=<x=  3, y=  3, z=  0>`
`pos=<x= -2, y= -4, z=  0>, vel=<x=  4, y=-13, z=  2>`
`pos=<x=-18, y= -7, z= 15>, vel=<x= -8, y=  2, z= -2>`
`pos=<x= -2, y= -1, z= -8>, vel=<x=  1, y=  8, z=  0>`

After 90 steps:
`pos=<x=-25, y= -1, z=  4>, vel=<x=  1, y= -3, z=  4>`
`pos=<x=  2, y= -9, z=  0>, vel=<x= -3, y= 13, z= -1>`
`pos=<x= 32, y= -8, z= 14>, vel=<x=  5, y= -4, z=  6>`
`pos=<x= -1, y= -2, z= -8>, vel=<x= -3, y= -6, z= -9>`

After 100 steps:
`pos=<x=  8, y=-12, z= -9>, vel=<x= -7, y=  3, z=  0>`
`pos=<x= 13, y= 16, z= -3>, vel=<x=  3, y=-11, z= -5>`
`pos=<x=-29, y=-11, z= -1>, vel=<x= -3, y=  7, z=  4>`
`pos=<x= 16, y=-13, z= 23>, vel=<x=  7, y=  1, z=  1>`

Energy after 100 steps:
`pot:  8 + 12 +  9 = 29;   kin: 7 +  3 + 0 = 10;   total: 29 * 10 = 290`
`pot: 13 + 16 +  3 = 32;   kin: 3 + 11 + 5 = 19;   total: 32 * 19 = 608`
`pot: 29 + 11 +  1 = 41;   kin: 3 +  7 + 4 = 14;   total: 41 * 14 = 574`
`pot: 16 + 13 + 23 = 52;   kin: 7 +  1 + 1 =  9;   total: 52 *  9 = 468`
`Sum of total energy: 290 + 608 + 574 + 468 = 1940`

What is the *total energy* in the system after simulating the moons given in your scan for *1000 steps*?

```{r custom functions}
parse_input <- function(.x) {
  .x <- 
    .x %>% 
    str_replace_all("<|>|.=|( )", "") %>% 
    read_csv(col_names = FALSE, col_types = cols(.default = "i"))
  
  names(.x) <- c("posx", "posy", "posz")
  
  .x <- mutate(.x, 
               moon = letters[1:4], 
               velx = 0L, 
               vely = 0L, 
               velz = 0L) %>% 
    relocate(moon)
  
  return(.x)
}
```


```{r tests}
test1 <- "<x=-1, y=0, z=2>
<x=2, y=-10, z=-7>
<x=4, y=-8, z=8>
<x=3, y=5, z=-1>"%>% 
  parse_input()

test2 <- "<x=-8, y=-10, z=0>
<x=5, y=5, z=10>
<x=2, y=-7, z=3>
<x=9, y=-8, z=-3>" %>% 
  parse_input()
```


```{r input}
input_raw <- read_file("Day12_input.txt")

input <- parse_input(input_raw)
```


# Puzzle 1 Solve

```{r puzzle 1}
puz1 <- input

velocity <- function(moon, others) {
  map_int(others, 
          ~ case_when(near(.x, moon) ~ 0L, 
                      .x > moon ~ 1L, 
                      .x < moon ~ -1L)) %>% 
    sum()
}

moon_step <- function(.df) {
  
  temp <- .df
  
  for (i in seq_along(.df$moon)) {
    
    moonx <- temp[i, "posx"][[1]]
    othersx <- temp[-i, "posx"][[1]]
    
    moony <- temp[i, "posy"][[1]]
    othersy <- temp[-i, "posy"][[1]]
    
    moonz <- temp[i, "posz"][[1]]
    othersz <- temp[-i, "posz"][[1]]
    
    .df[i, "velx"] <- temp[i, "velx"][[1]] + velocity(moonx, othersx)
    .df[i, "vely"] <- temp[i, "vely"][[1]] + velocity(moony, othersy)
    .df[i, "velz"] <- temp[i, "velz"][[1]] + velocity(moonz, othersz)
    
  }
  
  
  .df[, "posx"] <- .df[, "posx"][[1]] + .df[, "velx"][[1]]
  .df[, "posy"] <- .df[, "posy"][[1]] + .df[, "vely"][[1]]
  .df[, "posz"] <- .df[, "posz"][[1]] + .df[, "velz"][[1]]
  
  return(.df)
}

moon_eval <- function(.df_initial = puz1, steps) {
  
  i <- 0L
  
  while (i < steps) {
    
    .df_initial <- moon_step(.df = .df_initial)
    
    i <- i + 1L
  }
  
  return(.df_initial)
}


calc_energy <- function(.df) {
  .df[, -1] %>% # first column is moon name
    mutate(across(.fns = abs)) %>% 
    rowwise() %>% 
    mutate(pot = sum(c_across(starts_with("pos"))), 
           kin = sum(c_across(starts_with("vel"))), 
           tot = pot * kin) %>% 
    pluck("tot") %>% 
    sum()
}

moon_eval(steps = 1000) %>% 
  calc_energy()

```

After 1,000 steps, there is `14907` energy in my input.



# Puzzle 2 

All this drifting around in space makes you wonder about the nature of the universe. Does history really repeat itself? You're curious whether the moons will ever return to a previous state.

Determine the *number of steps* that must occur before all of the moons' *positions and velocities* exactly match a previous point in time.

For example, the first example above takes `2772` steps before they exactly match a previous point in time; it eventually returns to the initial state:

After 0 steps:
`pos=<x= -1, y=  0, z=  2>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  2, y=-10, z= -7>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  4, y= -8, z=  8>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  3, y=  5, z= -1>, vel=<x=  0, y=  0, z=  0>`

After 2770 steps:
`pos=<x=  2, y= -1, z=  1>, vel=<x= -3, y=  2, z=  2>`
`pos=<x=  3, y= -7, z= -4>, vel=<x=  2, y= -5, z= -6>`
`pos=<x=  1, y= -7, z=  5>, vel=<x=  0, y= -3, z=  6>`
`pos=<x=  2, y=  2, z=  0>, vel=<x=  1, y=  6, z= -2>`

After 2771 steps:
`pos=<x= -1, y=  0, z=  2>, vel=<x= -3, y=  1, z=  1>`
`pos=<x=  2, y=-10, z= -7>, vel=<x= -1, y= -3, z= -3>`
`pos=<x=  4, y= -8, z=  8>, vel=<x=  3, y= -1, z=  3>`
`pos=<x=  3, y=  5, z= -1>, vel=<x=  1, y=  3, z= -1>`

After 2772 steps:
`pos=<x= -1, y=  0, z=  2>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  2, y=-10, z= -7>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  4, y= -8, z=  8>, vel=<x=  0, y=  0, z=  0>`
`pos=<x=  3, y=  5, z= -1>, vel=<x=  0, y=  0, z=  0>`

Of course, the universe might last for a *very long time* before repeating. Here's a copy of the second example from above:

`<x=-8, y=-10, z=0>`
`<x=5, y=5, z=10>`
`<x=2, y=-7, z=3>`
`<x=9, y=-8, z=-3>`

This set of initial positions takes `4686774924` steps before it repeats a previous state! Clearly, you might need to find a more efficient way to simulate the universe.

*How many steps does it take* to reach the first state that exactly matches a previous state?

# Puzzle 2 Solve

So, based on how this problem is set up, each axis is independent of the others.  That is, the x-axis position and velocities only ever influence the x-axis values of the other moons.

Also, each output is dependent on the input so the first time any repeats occur should be the initial state. 

One way to solve this might be to figure out how soon each axis' positions and velocities repeat.  Then, once solved for each axis (x, y, z) the solution for the first time the all the positions and velocites repeat for the first time would be the least common multiple of the three axis'.

```{r puzzle2}
puz2 <- input

moon_step_puz2 <- function(.df) {
  
  temp <- .df
  
  for (i in seq_along(.df[[1]])) {
    
    moon <- temp[i, 1][[1]]
    others <- temp[-i, 1][[1]]
    
    .df[i, 2] <- temp[i, 2][[1]] + velocity(moon, others)
  }
  
  .df[, 1] <- .df[, 1][[1]] + .df[, 2][[1]]
  
  return(.df)
}

concat_eval <- function(.x) {.x %>% unlist() %>%  str_c(collapse = "")}


repeat_axis <- function(.df = puz2, axis) {
  
  .df <- select(.df, ends_with(axis)) 
  
  result <- concat_eval(.df)
  
  i <-  0L
  match  <-  FALSE
  
  while (match == FALSE) {
    
    i <- i + 1L
    
    .df <- moon_step_puz2(.df = .df)
    
    eval <- concat_eval(.df)
    
    if (sum(eval == result) > 0) {match <- TRUE}
  }
  
  return(i)
}


rep_x <- repeat_axis(axis = "x")
rep_y <- repeat_axis(axis = "y")
rep_z <- repeat_axis(axis = "z")


# cross-join and multiply to search for least common multiple
lcm <- function(a, b) {
  tibble(axis = c("a", "b"), 
         value = as.double(c(a, b))) %>% 
    crossing(n = 1:1e6) %>% 
    mutate(value = value * n) %>% 
    count(value) %>% 
    filter(n >= 2) %>% 
    arrange(value) %>% 
    head(1) %>% 
    pull(1)
}

# least common multiple of each of two variables, then lcm of those combinations.
xy <- lcm(rep_x, rep_y)
xz <- lcm(rep_x, rep_z)
xyz <- lcm(xy, xz)


```


`467081194429464` which is correct. (~467 trillion)


First got answer from Wolfram Alpha website, before coding solution in R above via lcm() custom function.  

Not the fastest answer for puzzle 2 and it takes a bit to run, particularly for the "repeat_axis" function, but it gets the correct answer for the puzzle!


